Density-dependent selection and the limits of relative fitness

### 2018-03-26 manuscript.tex
l

Here we argue that the simplified models of density-regulated growth mentioned above are misleading in some important ways. These ultimately derive from their failure to account for ``reproductive excesses'', that is, an excess of juveniles that are more fragile than their adult counterparts \citep{turner1968population}. Reproductive excesses have major repercussions for the interplay between density and selection, because they allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles. This concept is familiar to geneticists, as exemplified by the Wright-Fisher model, where each generation involves the production of an infinite number of zygotes, the death of all $N$ adults, followed by the sampling of $N$ of these zygotes to form the next generation of adults. Similarly, in the Moran model, there is always a juvenile ready to replace an adult death, and so $N$ remains constant.

The Wright-Fisher and Moran models have infinite reproductive excess --- population density $N$ remains constant no matter how strong the selection on juveniles. In reality,  reproductive excesses are finite, and the dynamics of density-dependent selection will depend on the magnitude of excess. Modeling finite excesses are not straightforward

Moreover, similar ideas featured prominently in early debates on the regulation of population density (e.g. \citealt{nicholson_1954}).

few attempts have been made to explicitly  It is not straightforward do so, because

\cite{nei1971fertility} proposed a model with an explicit reproductive excess as part of a discussion of Haldane's ``cost of selection'' \citep{haldane_1957}, but used an unusual model of competition which was defined for at most two different types. In more complex models with detailed age structure (as opposed to juvenile vs adult), it is usually assumed that the density of a ``critical age-group'' controls the population's response to crowding, but otherwise the view of density dependence is surprisingly similar to that found in the models mentioned above, which have no age structure \citep[Chap. 4]{charlesworth_1994}. For example, selection proceeds in the direction of increasing equilibrium density in the critical age-group \citep[pp. 148]{charlesworth_1994}. As a result, the extent to which reproductive excesses justify the use of constant relative finesses is unclear beyond heuristic verbal arguments.

>>>>

Here we argue that the simplified models of density-regulated growth mentioned above are misleading in some important ways. These ultimately derive from their failure to account for ``reproductive excesses'', that is, an excess of juveniles that are more fragile than their adult counterparts \citep{turner1968population}. This has major repercussions for the interplay between density and selection, because selection can be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles. Reproductive excesses featured prominently in early debates on the regulation of population density (e.g. \citealt{nicholson_1954}). They are also conceptually familiar in evolutionary biology, particularly in relation to Haldane's ``cost of selection'' \citep{haldane_1957,turner1968population}. Reproductive excesses even appear implicitly in foundational evolutionary-genetic models like the Wright-Fisher, where each non-overlapping generation involves the production of an infinite number of zygotes, a constant number $N$ of which are sampled to form the next generation of adults. Likewise, in the Moran model, there are always juveniles available to replace dead adults, and so $N$ remains constant. Nevertheless, few attempts have been made to study density-dependent selection in the presence of a reproductive excess.

More specifically, what is missing is the presence of a finite reproductive excess, rather than an infinite excess as in the Wright-Fisher.

the dynamics of both selection and density depend on the magnitude of a

Yet the properties of density-dependent selection in the presence of reproductive excesses

Wright-Fisher and Moran models both have inifinite excesses,

few attempts have been made to model the consequences of finite reproductive excesses,

 explicit models of finite reproductive excesses

Reproductive excesses are challenging to model in this context,

In reality, reproductive excesses are finite,

 Modeling density-dependent selection in the presence of finite reproductive excesses

Modeling finite excesses are not straightforward

\cite{nei1971fertility} proposed a model with an explicit reproductive excess as part of a , but used an unusual model of competition which was defined for at most two different types. In more complex models with detailed age structure (as opposed to juvenile vs adult), it is usually assumed that the density of a ``critical age-group'' controls the population's response to crowding, but otherwise the view of density dependence is surprisingly similar to that found in the models mentioned above, which have no age structure \citep[Chap. 4]{charlesworth_1994}. For example, selection proceeds in the direction of increasing equilibrium density in the critical age-group \citep[pp. 148]{charlesworth_1994}. As a result, the extent to which reproductive excesses justify the use of constant relative finesses is unclear beyond heuristic verbal arguments.
